Response to reviewers

### 2025-01-23 Manuscript/ecography_reponse_letter_2.tex
nearly done?

\textcolor{blue}{Lines 330 – 335/ Figure 4: I appreciate the clarification regarding the TPD metric.
However, I regret to say that comparing a metric built on four dimensions with one working on five dimensions does not seem appropriate, particularly given the aims of the manuscript.
Considering the amount of variation explained by the first four dimensions (Supplementary Figure 11) and the manuscript's focus on exploring metric performance with a real dataset rather than depicting bird diversity after extinction, standardizing the dimensionality across all metrics is necessary.
This approach would enable a fair comparison, even if it requires a slight compromise on the 95\% variation threshold, targeting approximately 90\% instead (as observed in the histogram for four dimensions, which aligns with the minimum number of dimensions recommended by Pigot et al. 2020).
This adjustment is particularly important, as metric performance may vary significantly across different dimensionalities (e.g., metrics evaluated across 2, 4, or 8 dimensions).
Standardizing to four dimensions would also ensure consistency with the dimensionality used in the simulated dataset and maintain uniformity throughout the analyses.
Regarding the 8-dimensional space, since it is not central to the analysis, I suggest removing the TPD approach for this case to avoid introducing unnecessary variability, while clearly justifying this decision as a limit of the package in the text.
Furthermore, upon focusing on Supplementary Figure 11, I have a concern.
Based on the description in the main text, I understand that the analyses are performed starting from a single trait space (comprising 118 points), where species are removed either randomly or based on actual extinction events, and metrics are then recalculated.
This approach seems appropriate for evaluating how species are distributed and how metric performance changes under different scenarios.
However, the figure comparing extant and pre-1500 species leads me to wonder whether the trait spaces were recalculated based on the subset of species following extinctions.
Am I correct in assuming that the analyses are based on a single trait space constructed with all 118 points?}

@@

\textit{-The \texttt{TPD} package currently only allows calculations for traitspaces up to four dimensions. This metric was exclude from traitspaces with more than four dimensions.}
% TODO: clarify the artificial extinctions in the birds dataset (yes, we always start from 118 points).
% TODO: double check if that's what the reviewer means.

@@

% TODO: say we're keeping Box 2 because.
% TODO: but also develop points they want.

>>>>

\textcolor{blue}{Lines 330 – 335/ Figure 4: I appreciate the clarification regarding the TPD metric.
However, I regret to say that comparing a metric built on four dimensions with one working on five dimensions does not seem appropriate, particularly given the aims of the manuscript.
Considering the amount of variation explained by the first four dimensions (Supplementary Figure 11) and the manuscript's focus on exploring metric performance with a real dataset rather than depicting bird diversity after extinction, standardizing the dimensionality across all metrics is necessary.
This approach would enable a fair comparison, even if it requires a slight compromise on the 95\% variation threshold, targeting approximately 90\% instead (as observed in the histogram for four dimensions, which aligns with the minimum number of dimensions recommended by Pigot et al. 2020).
This adjustment is particularly important, as metric performance may vary significantly across different dimensionalities (e.g., metrics evaluated across 2, 4, or 8 dimensions).
Standardizing to four dimensions would also ensure consistency with the dimensionality used in the simulated dataset and maintain uniformity throughout the analyses.
Regarding the 8-dimensional space, since it is not central to the analysis, I suggest removing the TPD approach for this case to avoid introducing unnecessary variability, while clearly justifying this decision as a limit of the package in the text.}

@@

\textit{-The \texttt{TPD} package currently only allows calculations for traitspaces up to four dimensions. This metric was exclude from traitspaces with more than four dimensions.}

\textcolor{blue}{Furthermore, upon focusing on Supplementary Figure 11, I have a concern.
Based on the description in the main text, I understand that the analyses are performed starting from a single trait space (comprising 118 points), where species are removed either randomly or based on actual extinction events, and metrics are then recalculated.
This approach seems appropriate for evaluating how species are distributed and how metric performance changes under different scenarios.
However, the figure comparing extant and pre-1500 species leads me to wonder whether the trait spaces were recalculated based on the subset of species following extinctions.
Am I correct in assuming that the analyses are based on a single trait space constructed with all 118 points?}

Yes, the empirical trait space is based on 118 observations split between three groups: 37 species present on the island today, 63 species that were present on island in 1500 and 18 species that went extinct before 1500.
We believe that the confusion here might come from the equivalence between species present (63 at 1500 and 37 in the present) and species absent (55 (46\%) at 1500 and 81 (68\%) in the present).
We have tried now to clarify this in the caption of the figure 3 and in the one from the supplementary figure concerning the dimensionality selection.

@@

We believe that this specific point conflicts with reviewers 1 (``The addition of the text box at the end is particularly appreciated, as it effectively highlights the key message of the manuscript'' - reviewer 1 above).
In this instance we thus prefer keeping the box 2 since we believe it might add value to at least some readers (and hopefully doesn't distract from the reading for other readers).
Furthermore, it is not entirely clear to us what the reviewer meant with the metrics relationship ``varying across spaces with different dimensionalities''.
Did they meant that spaces with different dimensionalities affect different metrics differently?
We did mention the effect of the curse in the methods section:

\textit{This curse changes the properties of space occupancy in a non-linear way depending on each observation's distribution and the number of dimensions.} lines @@@

We now also mention it in the discussion:

\textit{Note also that, as mentioned above, different dimensionalities can have different effects on the metrics, depending on both the traitspace and the metric nature \citep{bellman1957dynamic}.} Lines @@@

### 2025-01-23 Manuscript/main_ecography_R2.tex
nearly done?

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.4\textwidth]{Figures/results_per_metric_per_stressor_emp.pdf}
\caption{\scriptsize{\textbf{Empirical results:} the metrics and the scaled changes are measured in the same way as in figure 2 (green = richness, yellow = divergence, blue = regularity).
Plain circles represent the metrics for the species present on Hawaii before 1500 (historic data; 46\% removal) and fainted squares for the species currently present on Hawaii (extant data; 69\% removal).
The model results (including $R^{2}$ and t-statistics) are available in supplementary table 4.
}}
\label{Fig:empirical_results}
\end{figure}
\bigskip

@@

We tested 12 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
Our results show that different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).
Therefore, the choice of trait space occupancy metric is essential to accurately describe a pattern of interest.
We found that different metrics can capture changes in pattern relative to the null mechanisms in different ways depending on the approximated mechanism of interest.
For example, for the approximated equalising mechanism, most metrics capture a pattern clearly smaller than the one from the null mechanism (i.e. randomly removing elements has more effects on the metric than removing them following the mechanism).
However, some metrics (e.g. the Functional evenness based on the hypervolume) capture a pattern clearly greater than the one from the null mechanism.
This means that changes in metric values can be counter intuitive and should always been interpreted in the context of the metric, data and question at hand.

>>>>

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.4\textwidth]{Figures/results_per_metric_per_stressor_emp.pdf}
\caption{\scriptsize{\textbf{Empirical results:} the metrics and the scaled changes are measured in the same way as in figure 2 (green = richness, yellow = divergence, blue = regularity).
Plain circles represent the metrics for the species present on Hawaii at 1500 (historic data; 46\% removal - 63 species present) and fainted squares for the species currently present on Hawaii (extant data; 69\% removal - 37 species present).
The model results (including $R^{2}$ and t-statistics) are available in supplementary table 4.
}}
\label{Fig:empirical_results}
\end{figure}
\bigskip

@@

We tested 12 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
Our results show that different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).
Therefore, the choice of trait space occupancy metric is essential to accurately describe a pattern of interest.
We found that different metrics can capture changes in pattern relative to the null mechanisms in different ways depending on the approximated mechanism of interest.
For example, for the approximated equalising mechanism, most metrics capture a pattern clearly smaller than the one from the null mechanism (i.e. randomly removing elements has more effects on the metric than removing them following the mechanism).
However, some metrics (e.g. the Functional evenness based on the hypervolume) capture a pattern clearly greater than the one from the null mechanism.
This means that changes in metric values can be counter intuitive and should always been interpreted in the context of the metric, data and question at hand.
Note also that, as mentioned above, different dimensionalities can have different effects on the metrics, depending on both the traitspace and the metric nature \citep{bellman1957dynamic}.

### 2025-01-23 Manuscript/supplementaries.tex
added raw values
updated responses
nearly done?

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_8d.pdf}
\caption{\scriptsize{\textbf{Simulation results for 8 dimensions:} the y axes represent the different metrics tested (sorted by categories).
The different columns represent the different stressors. The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
}}
\label{Fig:simulation_results_8d}
\end{figure}
\bigskip

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_2d500.pdf}
\caption{\scriptsize{\textbf{Simulation results for 2 dimensions and 500 observations (instead of 200):} the y axes represent the different metrics tested (sorted by categories).
The different columns represent the different stressors. The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
}}
\label{Fig:simulation_results_2d500}
\end{figure}
\bigskip

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_2d1000.pdf}
\caption{\scriptsize{\textbf{Simulation results for 2 dimensions and 1000 observations (instead of 200):} the y axes represent the different metrics tested (sorted by categories).
The different columns represent the different stressors. The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
}}
\label{Fig:simulation_results_2d1000}
\end{figure}
\bigskip

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_4d500.pdf}
\caption{\scriptsize{\textbf{Simulation results for 4 dimensions and 500 observations (instead of 200):} the y axes represent the different metrics tested (sorted by categories).
The different columns represent the different stressors. The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
}}
\label{Fig:simulation_results_4d500}
\end{figure}
\bigskip

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_4d1000.pdf}
\caption{\scriptsize{\textbf{Simulation results for 2 dimensions and 1000 observations (instead of 200):} the y axes represent the different metrics tested (sorted by categories).
The different columns represent the different stressors. The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
}}
\label{Fig:simulation_results_4d1000}
\end{figure}
\bigskip

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{Figures/raw_results_per_metric_per_stressor_2d.pdf}
\caption{\scriptsize{\textbf{Raw metric scores for 8 dimensions:} the y axes represent the different metrics tested (sorted by categories).
The different columns represent the different stressors.
The x-axes represent the scaled metric scores.
The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
}
\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
}
\label{Fig:raw_results_8d}
\end{figure}
\bigskip

@@

\begin{figure}[!htbp]
\centering
   \includegraphics{Figures/empirical_data_dimensions.pdf}
\caption{\scriptsize{\textbf{Selection of the number of dimensions for the empirical data:} each histograms corresponds to the variance per dimension (dark grey) and the cumulative variance per dimension (light grey) for each group in the empirical data: the 37 currently endemic Hawaiian species (extant species) the 63 species present before 1500 (pre-1500 group). The plot shows that although 4 dimensions are enough to describe at least 95\% of the variance for the extant group, 5 dimensions are required for the pre-1500 group (following the method proposed in \cite{guillerme2023innovation}).
}}
\label{Fig:empirical_dimensions}
\end{figure}
\bigskip

>>>>

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_8d.pdf}
\caption{\scriptsize{\textbf{Simulation results for 8 dimensions:} the y axes represent the different metrics tested (sorted by categories).
The different columns represent the different stressors. The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
}}
\label{Fig:simulation_results_8d}
\end{figure}
\bigskip

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_2d500.pdf}
\caption{\scriptsize{\textbf{Simulation results for 2 dimensions and 500 observations (instead of 200):} the y axes represent the different metrics tested (sorted by categories).
The different columns represent the different stressors. The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
}}
\label{Fig:simulation_results_2d500}
\end{figure}
\bigskip

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_2d1000.pdf}
\caption{\scriptsize{\textbf{Simulation results for 2 dimensions and 1000 observations (instead of 200):} the y axes represent the different metrics tested (sorted by categories).
The different columns represent the different stressors. The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
}}
\label{Fig:simulation_results_2d1000}
\end{figure}
\bigskip

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_4d500.pdf}
\caption{\scriptsize{\textbf{Simulation results for 4 dimensions and 500 observations (instead of 200):} the y axes represent the different metrics tested (sorted by categories).
The different columns represent the different stressors. The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
}}
\label{Fig:simulation_results_4d500}
\end{figure}
\bigskip

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{Figures/results_per_metric_per_stressor_4d1000.pdf}
\caption{\scriptsize{\textbf{Simulation results for 2 dimensions and 1000 observations (instead of 200):} the y axes represent the different metrics tested (sorted by categories).
The different columns represent the different stressors. The x-axes represent the metric values centred on the random changes and scaled by the maximum value for each metric between the four stressors.
Negative and positive values signify that the metric score for the stressor of interest is lower/higher than the one from the random stressor.
The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
The grey line plots represent distributions of metric scores not clearly distinguishable from the random metric scores (paired t-test p value $> 0.05$).
Grey lines in the background across the distributions of different removal amounts represent the fitted linear model centred and scaled metric score $\sim$ amount of data removed and the value displayed is the adjusted $R^2$ from each of these models.
Dashed thin grey lines represent non-significant models (p value of slope or/and intercept $> 0.05$).
}}
\label{Fig:simulation_results_4d1000}
\end{figure}
\bigskip

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{Figures/raw_results_per_metric_per_stressor_8d.pdf}
\caption{\scriptsize{\textbf{Raw metric scores for 8 dimensions:} the y axes represent the different metrics tested (sorted by categories).
The different columns represent the different stressors.
The x-axes represent the scaled metric scores.
The dots represent the median metric value, the full line their 50\% confidence interval (CI) and the dashed line their 95\% CI.
The colours are here to visually separate the metrics by categories (blue = regularity, yellow = divergence, green = richness); the colour gradient within each row corresponds to a removal of respectively 80\%, 60\%, 40\% and 20\% of the data (from top to bottom).
}
\textbf{NOTE: for the \texttt{TPD} metrics, the metric is calculated using the 4 first dimensions only.}
}
\label{Fig:raw_results_8d}
\end{figure}
\bigskip

@@

\begin{figure}[!htbp]
\centering
   \includegraphics{Figures/empirical_data_dimensions.pdf}
\caption{\scriptsize{\textbf{Selection of the number of dimensions for the empirical data:} each histograms corresponds to the variance per dimension (dark grey) and the cumulative variance per dimension (light grey) for each group in the empirical data: the 37 currently endemic Hawaiian species (extant species) the 63 species present at 1500 (at_1500 group). The plot shows that although 4 dimensions are enough to describe at least 95\% of the variance for the extant group, 5 dimensions are required for the at_1500 group (following the method proposed in \cite{guillerme2023innovation}). Note that the whole trait space (light grey) contains 118 species. The 18 species extinct before 1500 are not contained in either of these groups.
}}
\label{Fig:empirical_dimensions}
\end{figure}
\bigskip
